---
title: "Basic grainsizeR Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic grainsizeR Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This vignette shows a complete sediment grain-size workflow with grainsizeR.
The examples use package data stored in `inst/extdata`.

For a more detailed discussion of table layouts versus measurement workflows,
including the dry-sieve and sieve + hydrometer example files, see the table
layouts and measurement workflows vignette.

```{r}
library(grainsizeR)
```

## Reading Long-Format Data

Long-format input has one row per sample and grain-size class.

```{r}
long_file <- system.file("extdata", "grain.long.csv", package = "grainsizeR")

gs <- read_gsd(
  long_file,
  format = "long",
  sample_col = "sample",
  size_col = "size",
  value_col = "proportion",
  size_unit = "mm",
  value_type = "proportion"
)

head(gs)
```

## Reading Wide-Format Data

Wide input stores size classes in rows and samples in columns. Terminal fine
rows such as `<0.0625` are supported and become open-ended fine classes.
G2Sd-style wide tables often store particle sizes in row names and samples in
columns. Convert row names to a size column and use `size_unit = "auto"` or
`size_unit = "um"` when importing micrometre-scale labels.

```{r}
wide_file <- system.file("extdata", "grain.wide.csv", package = "grainsizeR")

gs_wide <- read_gsd(
  wide_file,
  format = "wide",
  size_col = 1,
  size_unit = "mm",
  value_type = "percent"
)

head(gs_wide)
```

## The `gsd_tbl` Object

`gsd_tbl` stores retained percentages, finite class boundaries, cumulative
percent finer and coarser values, and open-ended class flags.

```{r}
is_gsd_tbl(gs)
validate_gsd_tbl(gs)
```

## Data Quality Diagnostics

`gs_diagnostics()` is useful after reading data and before calculating summary
tables. It reports retained-total checks, open-ended tails, unresolved D-values,
threshold interpolation limits, and fraction-scheme resolvability. The example
wide file is a dry-sieve dataset stored in a wide layout, so fine clay/silt
thresholds may be unresolved unless finer measurements or explicit
extrapolation are supplied.

```{r}
head(gs_diagnostics(gs_wide, output = "summary"))
head(gs_diagnostics(
  gs,
  d_values = c(5, 10, 50, 90, 95),
  fraction_schemes = c("wentworth_major", "usda_tt", "uk_ssew")
))
```

## Cumulative Curves and D-Values

`D_p` is the grain size at which `p` percent of the sample is finer.

```{r}
head(gs_cumulative(gs))
head(gs_d_values(gs, probs = c(10, 50, 90)))
```

## Arbitrary Threshold Interpolation

`gs_percent_finer()` estimates percent finer at arbitrary thresholds, including
thresholds that were not measured exactly, as long as the threshold is bracketed
by finite class boundaries. Terminal open-ended classes require explicit
extrapolation or unresolved-value handling.

```{r}
head(suppressWarnings(gs_percent_finer(
  gs,
  sizes = c(2, 20, 50, 60, 63),
  size_unit = "um",
  extrapolate = "warn_linear"
)))
```

## Grain-Size Indices

```{r}
head(suppressWarnings(gs_grain_size_indices(
  gs,
  extrapolate = "warn_linear"
)))
```

## Folk and Ward Statistics

```{r}
head(suppressWarnings(gs_folk_ward(
  gs,
  extrapolate = "warn_linear"
)))
```

## Moment Statistics and Open-Ended Classes

Moment statistics require an explicit open-end policy. `open_end =
"extend_phi"` estimates open-ended midpoints by extending adjacent intervals in
phi space.

```{r}
head(suppressWarnings(gs_moments(
  gs,
  open_end = "extend_phi"
)))
```

## Grain-Size Fractions and Particle-Size Systems

```{r}
head(gs_fractions_wide(gs, scheme = "wentworth_major"))
particle_size_systems()
```

## Creating a Report Table

`gs_parameters()` is the recommended way to create a compact grain-size summary
table for reporting. It can combine D-values, grain-size indices, Folk and Ward
statistics, moment statistics, and fractions in one row per sample.

```{r}
report_table <- suppressWarnings(gs_parameters(
  gs,
  parameters = c("d_values", "indices", "folk_ward", "fractions"),
  fraction_scheme = "gradistat",
  extrapolate = "warn_linear"
))

head(report_table)
```

`plot_gradistat_summary()` is for visual diagnostics. Standard R functions can
export tables when needed:

```{r, eval=FALSE}
write.csv(report_table, "grain_size_summary.csv", row.names = FALSE)
```

Keeping export in standard R workflows avoids tying grainsizeR to a specific
report format.

## Basic Plots

```{r, fig.width=6, fig.height=4}
plot_distribution(gs, sample = "S01")
```

```{r, fig.width=6, fig.height=4}
plot_cumulative(gs, sample = "S01", show_percentiles = c(10, 50, 90), extrapolate = "warn_linear")
```

```{r, fig.width=6, fig.height=4}
plot_fractions(gs, scheme = "wentworth_major")
```

`plot_gradistat_summary()` provides a single-sample diagnostic plot that
combines retained distribution, cumulative percent finer, D-value markers, and
summary statistics. It is inspired by common sediment grain-size reporting
needs, not copied from GRADISTAT software.

```{r, fig.width=7, fig.height=5}
suppressWarnings(plot_gradistat_summary(
  gs,
  sample_id = "S01",
  extrapolate = "warn_linear"
))
```

## Recommended Reporting Workflow

For reproducible reporting, state the input format, measurement units,
interpolation scale, open-end policy, fraction scheme, and whether any values
were extrapolated or unresolved.
