---
title: "Grain-Size Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grain-Size Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

This vignette shows a compact end-to-end grain-size workflow with grainsizeR.
It uses the example files installed with the package and keeps each step as an
ordinary R object so results can be checked, joined, plotted, and exported with
standard R tools.

This vignette uses preferred public functions in the main workflow. Short
aliases are available for interactive use, but the full function names are
easier to read in scripts and reports. CRAN readiness is not claimed here.

```{r}
library(grainsizeR)
```

## Reading Long and Wide Grain-Size Data

Long input has one row per sample and size class. Wide input has one size
column and one column per sample.

```{r}
long_file <- system.file("extdata", "grain.long.csv", package = "grainsizeR")
wide_file <- system.file("extdata", "grain.wide.csv", package = "grainsizeR")

gs <- read_gsd(
  long_file,
  format = "long",
  sample_col = "sample",
  size_col = "size",
  value_col = "proportion",
  size_unit = "mm",
  value_type = "proportion"
)

gs_wide <- read_gsd_wide(
  wide_file,
  size_col = 1,
  size_unit = "mm",
  value_type = "percent"
)

head(gs)
```

## Inspecting and Validating `gsd_tbl` Objects

`read_gsd()` and `read_gsd_wide()` return `gsd_tbl` objects. Diagnostics help
identify open-ended classes and threshold-resolution limits before running a
large analysis.

```{r}
is_gsd_tbl(gs)
head(gs_diagnostics(gs, output = "summary"))
```

## Cumulative Percentages

`gs_cumulative()` returns finite-boundary cumulative percentages by sample.

```{r}
head(gs_cumulative(gs))
```

## D-Values and GRADISTAT-Style D-Spread Descriptors

Use `gs_d_values()` for requested percentiles and `gs_d_spread()` for
GRADISTAT-style D-ratio and D-difference descriptors. Open-ended tails require
an explicit extrapolation choice when a requested value falls outside resolved
boundaries.

```{r}
head(suppressWarnings(gs_d_values(
  gs,
  probs = c(10, 50, 90),
  extrapolate = "warn_linear"
)))

head(suppressWarnings(gs_d_spread(
  gs,
  extrapolate = "warn_linear"
)))
```

## Folk and Ward Graphical Statistics

`gs_folk_ward()` calculates graphical statistics from percentile estimates.

```{r}
head(suppressWarnings(gs_folk_ward(
  gs,
  extrapolate = "warn_linear"
)))
```

## Moment Statistics

Moment statistics require explicit open-end handling. This example extends
terminal classes by one phi unit.

```{r}
head(suppressWarnings(gs_moments(
  gs,
  open_end = "extend_phi"
)))
```

## Modes and Sample Modality

`gs_modes()` reports ranked retained-class modes and an operational modality
label.

```{r}
head(gs_modes(gs))
```

## Fraction Summaries

Fraction summaries can be returned in long or wide form. The long form is
convenient for plotting and joins; the wide form is convenient for reports.
The dry-sieve wide example is useful for GRADISTAT-style gravel-sand-mud
summaries. The long example includes finer fractions and is used later for
USDA texture workflows.

```{r}
head(gs_fractions(gs, scheme = "wentworth_major"))
head(gs_fractions_wide(gs, scheme = "wentworth_major"))
head(gs_fractions_wide(gs_wide, scheme = "gradistat"))
```

## Descriptive Terms

`gs_describe_parameters()` adds GRADISTAT-style printout descriptors for
calculated Folk and Ward or moment statistics.

```{r}
head(suppressWarnings(gs_describe_parameters(gs)))
```

## Quality Flags

`gs_quality_flags()` records advisory flags for supplied sediment loss and
open-ended fine pan fractions.

```{r}
head(gs_quality_flags(
  gs,
  sediment_loss_percent = c(WN1_upper = 1.5, WN2_upper = 2.5)
))
```

## Distribution Plots

`plot_distribution()` returns a ggplot object and supports metric and phi axis
scales. The same function can overlay cumulative percent finer on the retained
size-class bars for a GRADISTAT-style combined display. The examples below use
the dry-sieve wide dataset so the plotted samples align with the README
GRADISTAT-style showcase.

```{r}
plot_distribution(gs_wide, sample_id = c("WN1_upper", "WN2_upper"))
plot_distribution(gs_wide, sample_id = c("WN1_upper", "WN2_upper"), cumulative = TRUE)
plot_distribution(gs_wide, x_scale = "phi", type = "line", sample_id = c("WN1_upper", "WN2_upper"))
```

## Cumulative Plots

`plot_cumulative()` can also add D-value markers.

```{r}
suppressWarnings(plot_cumulative(
  gs_wide,
  sample_id = c("WN1_upper", "WN2_upper"),
  show_percentiles = c(10, 50, 90),
  extrapolate = "warn_linear"
))
```

## Fraction Plots

`plot_fractions()` draws size-class percentage bars. For dry-sieve
GRADISTAT-style examples, use non-overlapping `Gravel`, `Sand`, and `Mud`
fractions. More detailed Wentworth-style classes are available with
`scheme = "wentworth_detailed"` when the input resolves those boundaries.

```{r}
plot_fractions(
  gs_wide,
  scheme = "gravel_sand_mud",
  sample_id = c("WN1_upper", "WN2_upper"),
  fill_palette = "YlOrBr"
)
```

## Building a Combined Analysis Table

`gs_parameters()` collects common outputs into a single table. More specialized
helpers remain available when you need a narrower result.

```{r}
combined <- suppressWarnings(gs_parameters(
  gs,
  parameters = c("d_values", "d_spread", "folk_ward", "modes", "descriptors", "quality"),
  d_values = c(10, 50, 90),
  extrapolate = "warn_linear",
  moments_open_end = "extend_phi"
))

head(combined)
```

## Notes on Open-Ended Tails and Extrapolation

Open-ended terminal classes are common in sediment data. grainsizeR does not
silently turn them into closed intervals. Functions that need unresolved
thresholds require an explicit extrapolation or open-end option, which makes
the assumption visible in the analysis script.
