---
title: "Method Validation and Numerical Assumptions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Method Validation and Numerical Assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

This vignette documents numerical conventions used by grainsizeR and explains
how the package validates them with synthetic examples and package example
data. It is a method-audit document, not a new scientific method.

Detailed interpretation of table layouts versus measurement workflows is
covered in the table layouts and measurement workflows vignette.

```{r}
library(grainsizeR)
```

## Example Data Used for Validation

The package includes two example files:

```{r}
long_file <- system.file("extdata", "grain.long.csv", package = "grainsizeR")
wide_file <- system.file("extdata", "grain.wide.csv", package = "grainsizeR")

gs_long <- read_gsd(
  long_file,
  format = "long",
  sample_col = "sample",
  size_col = "size",
  value_col = "proportion",
  size_unit = "mm",
  value_type = "proportion"
)

gs_wide <- read_gsd(
  wide_file,
  format = "wide",
  size_col = 1,
  size_unit = "mm",
  value_type = "percent"
)
```

`grain.long.csv` is a sieve + hydrometer example stored as a tidy long table
and has finer particle-size resolution where hydrometer data were added.
`grain.wide.csv` is a dry-sieve example stored as a multi-sample wide table and
is useful for coarse gravel/sand/mud summaries.

## Input-Bin Convention

Input rows describe retained material in grain-size classes. A numeric label is
used as a class boundary. Terminal coarse and fine classes can be open-ended.
Open-ended classes are retained in the data structure, but their missing
boundary is not silently replaced by zero or infinity for interpolation.

## Long Versus Wide Example Data

The long and wide examples contain the same sample IDs. Here, the file names
describe table layout, not measurement method. Coarse Wentworth-style gravel,
sand, and mud summaries agree between the two files because both represent the
same aggregate behavior at 62.5 um. Clay/silt-level schemes such as USDA, ISSS,
and UK SSEW require thresholds such as 2, 20, 50, 60, or 63 um; these are
better resolved when finite measured boundaries bracket those thresholds.

Instrument outputs such as laser cumulative percent-finer curves may require
preprocessing before import. `read_gsd()` expects retained proportions or
retained percentages by size class, not cumulative percent passing.

## Percent-Finer Convention

grainsizeR uses cumulative percent finer. At a finite boundary, percent finer
is the retained percent below that boundary. This convention is used by
`gs_cumulative()`, `gs_percent_finer()`, and `gs_d_values()`.

## D-Values

`D_p` is the grain size at which `p` percent of the sample is finer. For
example, `D50` is the median grain size by the percent-finer convention.

```{r}
gs_d_values(subset(gs_long, sample_id == "S01"), probs = c(10, 50, 90), extrapolate = "warn_linear")
```

## Threshold Interpolation

`gs_percent_finer()` estimates percent finer at arbitrary thresholds by
interpolating the cumulative percent-finer curve between finite boundaries.
Thresholds do not need to be measured boundaries. This is how thresholds such
as 2, 20, 50, 60, and 63 um are handled when they are bracketed by finite
class boundaries.

```{r}
suppressWarnings(gs_percent_finer(
  subset(gs_long, sample_id == "S01"),
  sizes = c(20, 50, 60, 63),
  size_unit = "um",
  extrapolate = "warn_linear"
))
```

## Open-Ended Terminal Classes

Terminal open-ended classes are not silently treated as bounded intervals. For
example, a final class such as `<0.0625 mm` is not treated as `[0, 62.5]`.
When a requested D-value or threshold falls inside an open-ended class,
`extrapolate = "error"` reports the unresolved value. Users can explicitly set
`extrapolate = "warn_linear"` when extrapolation is acceptable for their
workflow.

`gs_diagnostics()` can be used before summary calculations to identify samples
with unresolved D-values, unresolved clay/silt thresholds, open-ended terminal
classes, or fraction schemes that require finer measurements.

```{r}
gs_diagnostics(subset(gs_wide, sample_id == "S01"), output = "summary")
```

## Fraction Schemes

Fraction schemes use `gs_percent_finer()` internally at scheme boundaries.
Coarse schemes such as `wentworth_major` can be robust when their thresholds
are resolved. Texture fraction schemes with fine thresholds should use input
data that bracket those thresholds with finite boundaries, regardless of
whether the table is long or wide.

```{r}
head(gs_fractions_wide(gs_long, scheme = "wentworth_major"))
head(gs_fractions_wide(gs_wide, scheme = "wentworth_major"))
```

## Folk and Ward Statistics

Folk and Ward statistics are calculated from boundary-interpolated D-values.
If a required D-value falls outside the finite boundary curve, users must
choose whether to error or explicitly extrapolate.

```{r}
suppressWarnings(gs_folk_ward(
  subset(gs_long, sample_id == "S01"),
  extrapolate = "warn_linear"
))
```

## Moment Statistics

Moment statistics require explicit open-end handling. `open_end = "error"`
prevents silent assumptions. `open_end = "extend_phi"` estimates open-ended
midpoints by extending adjacent phi intervals, and `open_end = "omit"` omits
open-ended retained classes.

```{r}
suppressWarnings(gs_moments(
  subset(gs_long, sample_id == "S01"),
  open_end = "extend_phi"
))
```

## Summary-Table Consistency

`gs_parameters()` is a reporting convenience wrapper around lower-level
functions. It does not introduce new calculations; tests compare its output
against `gs_d_values()`, `gs_grain_size_indices()`, `gs_folk_ward()`, and
`gs_fractions_wide()`.

```{r}
sample_id <- "S01"
summary <- suppressWarnings(gs_parameters(
  subset(gs_long, sample_id == sample_id),
  parameters = c("d_values", "indices", "folk_ward", "fractions"),
  fraction_scheme = "gradistat",
  extrapolate = "warn_linear"
))
summary
```

## What This Package Does Not Do

grainsizeR does not implement civil-engineering classification systems. Built-in
official texture polygon datasets are not bundled yet. User-supplied texture
polygons are supported, and source-audit scaffolds document how future
official polygon datasets should be reconstructed and reviewed.
