---
title: "Replacing GRADISTAT and G2Sd Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replacing GRADISTAT and G2Sd Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Purpose

This vignette shows how grainsizeR can be used as an R-native functional
replacement workflow for many GRADISTAT and G2Sd-style grain-size analysis
tasks. It is not an Excel visual clone and does not claim byte-for-byte parity
with workbook printouts.

```{r}
library(grainsizeR)
```

## What GRADISTAT and G2Sd Workflows Usually Provide

Typical workflows include importing retained grain-size data, calculating
D-values and graphical statistics, describing sediment parameters, checking
quality cautions, classifying texture, composing sediment names, and producing
distribution, cumulative, fraction, and texture ternary plots.

## grainsizeR Equivalent Functions

| GRADISTAT / G2Sd output | grainsizeR function | Notes |
|---|---|---|
| Retained grain-size input | `read_gsd()`, `read_gsd_wide()` | Long and wide retained tables are supported. |
| D-values | `gs_d_values()` | Percentile grain sizes with explicit extrapolation behavior. |
| D-ratio and D-difference descriptors | `gs_d_spread()` | GRADISTAT-style spread descriptors. |
| Folk and Ward statistics | `gs_folk_ward()` | Graphical statistics in R tabular output. |
| Moment statistics | `gs_moments()` | Explicit open-end handling is required. |
| Modes and modality | `gs_modes()` | Ranked retained-class modes and sample modality. |
| Fraction percentages | `gs_fractions()`, `gs_fractions_wide()` | Built-in particle-size schemes. |
| Descriptive terms | `gs_describe_parameters()` | GRADISTAT-style printout descriptors. |
| Quality cautions | `gs_quality_flags()` | Sediment loss and open fine-pan advisories. |
| Summary table | `gs_parameters()` | Combined R table for reporting. |
| Distribution plot | `plot_distribution()` | ggplot output with metric and phi scales. |
| Cumulative curve | `plot_cumulative()` | ggplot output with optional D-value markers. |
| Fraction plot | `plot_fractions()` | ggplot stacked bars by sample. |
| Texture classification | `classify_texture()` | USDA and GRADISTAT rule paths plus user polygons. |
| Sediment names | `gs_gradistat_sediment_name()` | GRADISTAT-style composition from texture classes. |
| Texture ternary plots | `plot_texture_ternary()` | Preferred terminology-correct alias; `plot_texture_triangle()` remains available for compatibility. |

Short aliases such as `gs_fw57()`, `gs_frac()`, `gs_diag()`, and `gs_qc()` are
available for interactive work. The full names in the table remain the clearest
choices for reproducible scripts and method descriptions.

## Input Data

```{r}
long_file <- system.file("extdata", "grain.long.csv", package = "grainsizeR")
wide_file <- system.file("extdata", "grain.wide.csv", package = "grainsizeR")

gs <- read_gsd(
  long_file,
  format = "long",
  sample_col = "sample",
  size_col = "size",
  value_col = "proportion",
  size_unit = "mm",
  value_type = "proportion"
)

gs_wide <- read_gsd(wide_file, format = "wide")

head(gs)
```

The wide dry-sieve example is used below for the GRADISTAT-style
gravel-sand-mud workflow. The long example includes finer fractions and is used
for USDA texture examples. Open-ended tails are not silently extrapolated; calls
that need extrapolation use `extrapolate = "warn_linear"` explicitly.

## Summary Statistics

```{r}
head(suppressWarnings(gs_parameters(
  gs,
  parameters = c("d_values", "indices", "folk_ward", "fractions"),
  d_values = c(10, 50, 90),
  fraction_scheme = "gradistat",
  extrapolate = "warn_linear"
)))
```

## D-Values and Spread Descriptors

```{r}
head(suppressWarnings(gs_d_values(gs, probs = c(10, 50, 90), extrapolate = "warn_linear")))
head(suppressWarnings(gs_d_spread(gs, extrapolate = "warn_linear")))
```

## Modes and Modality

```{r}
head(gs_modes(gs))
```

## Descriptive Terms and Quality Flags

```{r}
head(suppressWarnings(gs_describe_parameters(gs)))
head(gs_quality_flags(gs, sediment_loss_percent = c(WN1_upper = 1.2, WN2_upper = 2.4)))
```

## Texture Classification

USDA major texture classification uses sand, silt, and clay percentages. USDA
sand-size modifier subclasses remain future work.

```{r}
usda_fractions <- suppressWarnings(gs_fractions_wide(
  gs,
  scheme = "usda_tt",
  extrapolate = "warn_linear"
))

usda_samples <- data.frame(
  sample_id = usda_fractions$sample_id,
  sand = usda_fractions$sand_percent,
  silt = usda_fractions$silt_percent,
  clay = usda_fractions$clay_percent
)
usda_components <- c("sand", "silt", "clay")
usda_samples <- usda_samples[
  stats::complete.cases(usda_samples[usda_components]) &
    rowSums(usda_samples[usda_components] >= 0 & usda_samples[usda_components] <= 100) == 3 &
    abs(rowSums(usda_samples[usda_components]) - 100) < 1e-6,
]

head(classify_texture(usda_samples, scheme = "usda_tt", method = "rules"))
```

GRADISTAT texture classification supports the gravel-sand-mud and sand-silt-clay
no-gravel bases.

```{r}
gradistat_fractions <- suppressWarnings(gs_fractions_wide(gs_wide, scheme = "gravel_sand_mud"))

gsm <- data.frame(
  sample_id = gradistat_fractions$sample_id,
  gravel = gradistat_fractions$gravel_percent,
  sand = gradistat_fractions$sand_percent,
  mud = gradistat_fractions$mud_percent
)
gsm <- gsm[stats::complete.cases(gsm[c("gravel", "sand", "mud")]), ]

ssc <- data.frame(
  sample_id = c("A", "B", "C"),
  sand = c(95, 60, 20),
  silt = c(3, 30, 60),
  clay = c(2, 10, 20)
)

gsm_classified <- classify_texture(
  head(gsm, 6),
  scheme = "gradistat",
  method = "rules",
  basis = "gravel_sand_mud",
  include_sediment_name = TRUE
)

ssc_classified <- classify_texture(
  ssc,
  scheme = "gradistat",
  method = "rules",
  basis = "sand_silt_clay_no_gravel"
)

gsm_classified
ssc_classified
```

## Sediment Names

```{r}
gs_gradistat_sediment_name(gsm_classified)
```

## Distribution and Cumulative Plots

Metric distribution and cumulative plots use particle size in millimetres on a
log-scaled x-axis by default, with major breaks at 0.001, 0.01, 0.1, 1, and
10 mm. Distribution bars are centered at particle-size classes. Use
`particle_unit = "um"` for micrometre axes. They show one sample at a time; loop
over samples or arrange returned plots externally for multi-sample figures.
Lower open-ended classes are displayed at 0.0015 mm, or 1.5 um, for plotting only;
calculations are unchanged.

```{r}
plot_distribution(gs_wide, sample_id = "WN1_upper", cumulative = TRUE)
suppressWarnings(plot_cumulative(
  gs_wide,
  sample_id = "WN1_upper",
  show_percentiles = c(10, 50, 90),
  extrapolate = "warn_linear"
))
plot_fractions(
  gs_wide,
  scheme = "gravel_sand_mud",
  sample_id = c("WN1_upper", "WN2_upper"),
  fill_palette = "YlOrBr"
)
```

## Texture Ternary Plots

GRADISTAT gravel-sand-mud ternary plots place `Gravel` at the top, `Mud` at the
lower-left apex, and `Sand` at the lower-right apex. The plotting functions draw
ternary guides for gravel percentage and sand/mud ratio directly on the diagram
and suppress ordinary Cartesian x/y axes.

```{r}
plot_texture_ternary(
  gsm_classified,
  scheme = "gradistat",
  basis = "gravel_sand_mud",
  point_id = "sample_id"
)

plot_texture_ternary(
  ssc_classified,
  scheme = "gradistat",
  basis = "sand_silt_clay_no_gravel",
  point_id = "sample_id"
)
```

## What Differs From Excel-Based GRADISTAT

grainsizeR returns R objects, not fixed spreadsheet worksheets. Plots are
ggplot objects and can be styled or combined with ordinary R tools. The package
does not copy GRADISTAT VBA code, chart objects, or workbook printout layouts.

## What Is Not Claimed

This vignette does not claim full Excel visual parity, byte-for-byte output
identity, complete modified Udden-Wentworth subclass parity, or a CRAN release
claim. It demonstrates a functional replacement workflow for the implemented
package scope.

## Remaining Future Work

Future work includes a separate CRAN-specific audit before any CRAN submission
and deferred features such as USDA sand-size modifier subclasses, additional
texture systems, and civil-engineering classifications if they are scoped
later.

## Reproducibility Advantages in R

An R-native workflow keeps import choices, extrapolation assumptions,
classification settings, plots, and output tables in a script. That makes the
analysis easier to rerun, review, and version-control than a manual spreadsheet
workflow.
