---
title: "Table layouts and measurement workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Table layouts and measurement workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

This vignette separates three concepts that are easy to mix up:

1. Measurement method, such as dry sieve, sieve + hydrometer, pipette, laser
   diffraction, or another particle-size technique.
2. Table layout, such as a single-sample table, a multi-sample wide table, or a
   tidy long table.
3. Data type, such as retained percent by class, cumulative percent finer,
   frequency or density output, raw counts, or instrument-specific output.

grainsizeR currently reads retained proportions or retained percentages by
size class. It does not infer laboratory method from file shape alone.

```{r}
library(grainsizeR)
```

## Table Layout Is Not Measurement Method

`format = "long"` and `format = "wide"` describe how the file is arranged.
They do not identify the laboratory or instrument method. A wide table can
represent dry-sieve, hydrometer, pipette, laser-diffraction, or other
particle-size data. A long table can also represent any of those methods.

grainsizeR interprets size classes and retained proportions or percentages. It
does not automatically understand every instrument-specific export format.
Laser outputs are suitable when they are converted or exported as retained
size-class values. If a laser instrument exports cumulative percent finer or
cumulative percent passing, the cumulative curve must be converted to retained
bin increments before using `read_gsd()`.

## Single-Sample and Multi-Sample Layouts

A single-sample table commonly has one size column and one value column. A
multi-sample wide table commonly has one size column and multiple sample
columns. GRADISTAT-style single-sample input is conceptually similar to one
sample's size-value table. GRADISTAT-style multi-sample input is conceptually
similar to grainsizeR's wide-table input, with size classes as rows and samples
as columns.

grainsizeR's tidy long format is often more convenient for batch analysis,
plotting, joining metadata, and reproducible workflows. This comparison is
about data organization only; it does not copy GRADISTAT input templates or
workbook logic.

## Example Datasets

`grain.wide.csv` is a dry-sieve example stored as a multi-sample wide table. It
has coarser particle-size resolution and a terminal fine open-ended class.

`grain.long.csv` is a sieve + hydrometer example stored as a tidy long table.
It has finer particle-size resolution for samples where hydrometer data were
added. Hydrometer measurements were added only when the fraction finer than
approximately 0.063 mm exceeded 10%.

Because of that workflow rule, some samples may still be unable to resolve
fine-end D-values, clay thresholds, or other fine-end thresholds without
explicit extrapolation.

Pipette-method and laser-diffraction data can also be imported as wide or long
tables when they are arranged as size-class retained percentages or
proportions. Suitability for clay/silt/sand fractions depends on whether the
required thresholds are resolved by finite measured size boundaries, not on
whether the input file is wide or long.

## Dry-Sieve Example in a Wide Table

```{r}
wide_file <- system.file("extdata", "grain.wide.csv", package = "grainsizeR")

gs_wide <- read_gsd(
  wide_file,
  format = "wide",
  size_col = 1,
  size_unit = "mm",
  value_type = "percent"
)

head(gs_diagnostics(gs_wide, output = "summary"))
```

This example is appropriate for the GRADISTAT-compatible `gravel_sand_mud`
summary because the bundled dry-sieve files use a 0.063 mm terminal fine
boundary. Clay and silt thresholds such as 2, 20, 50, or 60 um may not be
resolvable from data with only a terminal open fine class. Open fine tails
should not be silently treated as bounded intervals. `wentworth_major` remains
available as a strict Wentworth / phi-scale alternative with the 62.5 um
sand/mud boundary; it is not an alias of `gravel_sand_mud`.

```{r}
head(gs_fractions_wide(gs_wide, scheme = "gravel_sand_mud"))

head(suppressWarnings(gs_parameters(
  gs_wide,
  parameters = c("d_values", "indices", "folk_ward", "fractions"),
  fraction_scheme = "gravel_sand_mud",
  extrapolate = "warn_linear"
)))
```

## Sieve + Hydrometer Example in a Long Table

```{r}
long_file <- system.file("extdata", "grain.long.csv", package = "grainsizeR")

gs_long <- read_gsd(
  long_file,
  format = "long",
  sample_col = "sample",
  size_col = "size",
  value_col = "proportion",
  size_unit = "mm",
  value_type = "proportion"
)

head(gs_diagnostics(gs_long, output = "summary"))
```

Sieve + hydrometer measurements are often preferred for clay/silt/sand texture
fractions because they can provide finite boundaries in the fine tail. In this
example, hydrometer was only added for samples with enough fine material, so
not every sample necessarily resolves all fine-end D-values. Unresolved D5 or
D95 values should be reported rather than hidden.

```{r}
head(suppressWarnings(gs_fractions_wide(gs_long, scheme = "usda")))
head(suppressWarnings(gs_fractions_wide(gs_long, scheme = "uk_ssew")))
```

## Why Some Fine-End D-Values Remain Unresolved

Open-ended terminal classes do not provide finite lower or upper boundaries.
If D5 falls inside an open fine tail, or D95 falls inside an open coarse tail,
the default behavior is to report that the value is unresolved. Users may
explicitly request extrapolation, but the extrapolation decision should be
documented.

## Diagnostics Before Analysis

Use diagnostics before clay/silt/sand fractions, D-values near open tails, or
texture classification.

```{r}
sample_id <- "S01"

head(gs_diagnostics(
  gs_long,
  d_values = c(5, 10, 50, 90, 95),
  fraction_schemes = c("wentworth_major", "usda", "uk_ssew")
))
```

## Coarse Fractions From the Dry-Sieve Example

For the dry-sieve example, `gravel_sand_mud` is usually the first coarse
fraction summary to inspect. It uses the GRADISTAT-compatible 63 um sand/mud
boundary. Use `wentworth_major` when a strict Wentworth / phi-scale 62.5 um
boundary is required.

```{r}
head(gs_fractions_wide(gs_wide, scheme = "gravel_sand_mud"))
```

## Clay/Silt/Sand Fractions From Finer-Resolution Data

Texture-fraction workflows require sufficient fine-resolution data to resolve
the relevant boundaries.

```{r}
head(suppressWarnings(gs_fractions_wide(gs_long, scheme = "usda")))
head(suppressWarnings(gs_fractions_wide(gs_long, scheme = "uk_ssew")))
```

## Summary Tables

`gs_parameters()` creates compact reporting tables from lower-level functions.

```{r}
head(suppressWarnings(gs_parameters(
  gs_long[gs_long$sample_id == sample_id, ],
  parameters = c("d_values", "indices", "folk_ward", "fractions"),
  fraction_scheme = "gradistat",
  extrapolate = "warn_linear"
)))
```

GRADISTAT-style D-spread descriptors, retained-class modes, printout
descriptors, and quality flags can be requested explicitly. D-spread ratios and
differences are metric descriptors, modal classes are ranked by retained class
percentage rather than estimated as continuous density peaks, and quality flags
are advisory checks rather than calculation blockers.

```{r}
head(suppressWarnings(gs_d_spread(
  gs_long[gs_long$sample_id == sample_id, ],
  extrapolate = "warn_linear"
)))

head(gs_modes(gs_long[gs_long$sample_id == sample_id, ]))

head(suppressWarnings(gs_parameters(
  gs_long[gs_long$sample_id == sample_id, ],
  parameters = c("d_spread", "modes", "descriptors", "quality"),
  extrapolate = "warn_linear"
)))
```

## GRADISTAT-Inspired Summary Plots

`plot_gradistat_summary()` creates a one-sample diagnostic/report plot. It does
not export files; users can save ggplot objects with standard R tools.

```{r, eval=FALSE}
plot_gradistat_summary(
  gs_long,
  sample_id = sample_id,
  extrapolate = "warn_linear"
)
```

## Recommended Workflow

1. Read data.
2. Validate the `gsd_tbl`.
3. Run `gs_diagnostics()`.
4. Decide whether the available measurements and table layout are appropriate
   for the target analysis.
5. Compute D-values, fractions, statistics, and summary tables.
6. Make diagnostic or report plots.
7. For future texture classification, use only data with resolved
   clay/silt/sand thresholds or document extrapolation explicitly.

## What This Vignette Does Not Do

This vignette does not add official texture polygon datasets. It does not
perform texture classification using built-in official polygons. It does not
implement civil-engineering classification systems. It does not silently fill
missing fine-tail information. It does not implement laser-specific import.
