---
title: "User-Supplied Texture Polygons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-Supplied Texture Polygons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Particle-size fraction schemes define boundaries such as clay, silt, sand, and
gravel limits. Texture class polygons define regions in a ternary diagram. The
two are related, but they are not the same data object.

grainsizeR includes particle-size fraction schemes and a framework for
user-supplied texture polygons. Official built-in polygon datasets are not
included yet because they need to be compiled from original official or
academic sources with citations.

```{r}
library(grainsizeR)

texture_polygon_sources()
```

USDA 12-class major texture ternary classification is available without a
bundled polygon dataset by using the validated internal rule path through
`classify_texture()`.

```{r}
samples <- data.frame(
  sample_id = c("A", "B", "C"),
  sand = c(90, 40, 20),
  silt = c(5, 40, 20),
  clay = c(5, 20, 60)
)

classify_texture(samples, scheme = "usda", method = "rules")
```

This USDA rule path covers only the 12 major classes and does not include
sand-size modifier subclasses. User-supplied polygon classification remains
the generic workflow for other texture systems.

## Creating a Polygon Data Frame

`texture_polygon_template()` shows the required columns. The synthetic example
below defines a single class covering the full ternary triangle. It is not an
official texture classification system.

```{r}
texture_polygon_template()

polygons <- data.frame(
  scheme = "synthetic_ternary",
  class_id = "all",
  class_name = "Synthetic full ternary area",
  vertex_id = 1:3,
  left = c(100, 0, 0),
  right = c(0, 100, 0),
  top = c(0, 0, 100),
  left_component = "sand",
  right_component = "silt",
  top_component = "clay",
  reference_id = NA_character_,
  reference = NA_character_
)

polygons <- validate_texture_polygons(polygons)
polygons
```

## Classifying Samples

This synthetic grain-size table has finite boundaries at the USDA-style sand,
silt, and clay thresholds used by the example polygon axes.

```{r}
synthetic <- data.frame(
  sample_id = rep(c("A", "B"), each = 4),
  size_mm = rep(c(2, 0.05, 0.002, 0.001), 2),
  retained = c(10, 40, 30, 20, 5, 20, 35, 40)
)

synthetic_gs <- as_gsd_tbl(
  synthetic,
  sample_id,
  size_mm,
  retained,
  value_type = "percent"
)

classify_texture(
  synthetic_gs,
  polygons = polygons,
  scheme = "synthetic_ternary"
)
```

## Plotting User Polygons

`plot_texture_ternary()` can draw sample points and user-supplied polygons.
`plot_texture_triangle()` remains available as a compatibility function name.

```{r, fig.width=6, fig.height=4}
plot_texture_ternary(
  synthetic_gs,
  polygons = polygons,
  scheme = "synthetic_ternary"
)
```
