---
title: "Texture Classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Texture Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`classify_texture()` provides two texture-classification paths:

- USDA 12-class major texture classification from sand, silt, and clay
  percentages.
- GRADISTAT-style rule classification for gravel-sand-mud textural groups and
  no-gravel sand-silt-clay mini texture classes.
- Generic classification against user-supplied ternary texture polygons.

The USDA path is rule-based. It does not require bundled USDA polygon data,
and no USDA polygon dataset is included as package data.

The GRADISTAT path is also rule-based. It uses source-grounded decision tables
from the user-provided GRADISTAT v8 workbook and the textural-output context
described by Blott and Pye (2001). Sediment-name composition is available as a
separate step or optional output. GRADISTAT ternary plotting is available
through `plot_texture_ternary()`.

```{r}
library(grainsizeR)
```

```{r}
long_file <- system.file("extdata", "grain.long.csv", package = "grainsizeR")
wide_file <- system.file("extdata", "grain.wide.csv", package = "grainsizeR")

long_gs <- read_gsd(long_file)
wide_gs <- read_gsd(wide_file, format = "wide")
```

## USDA Major Texture Classification

Use `scheme = "usda_tt"` and `method = "rules"` for the validated USDA
major-class workflow.

```{r}
samples <- data.frame(
  sample_id = c("A", "B", "C"),
  sand = c(85, 40, 20),
  silt = c(10, 40, 20),
  clay = c(5, 20, 60)
)

classify_texture(samples, scheme = "usda_tt", method = "rules")
```

USDA texture ternary plotting is available through `plot_texture_ternary()`.
The stable `plot_texture_triangle()` function name remains available for
compatibility. The bundled long-format example includes finer fractions and can
be used for the USDA texture workflow. Some open-ended tails require explicit
extrapolation for resolved sand, silt, and clay percentages; this example makes
that assumption visible. USDA ternary plots draw the 12 major-class boundaries
from the package's independently expressed rule logic and annotate ternary
percentage axes directly, without ordinary Cartesian x/y tick labels. USDA
ternary plots place `percent sand`, `percent silt`, and `percent clay` outside
the corresponding triangle sides.

```{r}
usda_fractions <- suppressWarnings(gs_fractions_wide(
  long_gs,
  scheme = "usda_tt",
  normalize = "fine_earth",
  extrapolate = "warn_linear"
))

usda_example <- data.frame(
  sample_id = usda_fractions$sample_id,
  sand = usda_fractions$sand_percent,
  silt = usda_fractions$silt_percent,
  clay = usda_fractions$clay_percent
)
usda_components <- c("sand", "silt", "clay")
usda_example <- usda_example[
  stats::complete.cases(usda_example[usda_components]) &
    rowSums(usda_example[usda_components] >= 0 & usda_example[usda_components] <= 100) == 3 &
    abs(rowSums(usda_example[usda_components]) - 100) < 1e-6,
]

head(classify_texture(usda_example, scheme = "usda_tt", method = "rules"))

plot_texture_ternary(usda_example, scheme = "usda_tt", labels = FALSE)
```

Input percentages must be numeric, finite, between 0 and 100, and sum to
approximately 100. `classify_texture()` does not silently normalize invalid
sand, silt, and clay sums.

Data frames with ternary `left`, `right`, and `top` columns are also accepted
for USDA rules. They are interpreted as `left = sand`, `right = silt`, and
`top = clay`.

## Method Selection

`method = "auto"` chooses the USDA rule path when `scheme = "usda_tt"` and no
texture polygons are supplied. It also chooses GRADISTAT rules when
`scheme = "gradistat"` and no texture polygons are supplied.

```{r}
classify_texture(samples, scheme = "usda_tt", method = "auto")
```

`method = "rules"` supports `scheme = "usda_tt"` and `scheme = "gradistat"`.
Other schemes require user-supplied polygons because grainsizeR does not bundle
built-in texture polygon datasets yet.

`method = "polygon"` uses the generic user-supplied polygon workflow.

## GRADISTAT Texture Classification

Use `scheme = "gradistat"` and `method = "rules"` for GRADISTAT-style texture
classification. The `gravel_sand_mud` basis requires `gravel`, `sand`, and
`mud` percentage columns. The bundled dry-sieve wide example supports this
GRADISTAT-style gravel-sand-mud workflow.

```{r}
wide_gradistat <- suppressWarnings(gs_fractions_wide(wide_gs, scheme = "gravel_sand_mud"))

gsm <- data.frame(
  sample_id = wide_gradistat$sample_id,
  gravel = wide_gradistat$gravel_percent,
  sand = wide_gradistat$sand_percent,
  mud = wide_gradistat$mud_percent
)
gsm <- gsm[stats::complete.cases(gsm[c("gravel", "sand", "mud")]), ]

classify_texture(
  head(gsm, 6),
  scheme = "gradistat",
  method = "rules",
  basis = "gravel_sand_mud",
  include_sediment_name = TRUE
)
```

The `sand_silt_clay_no_gravel` basis requires `sand`, `silt`, and `clay`
percentage columns.

```{r}
ssc <- data.frame(
  sample_id = c("A", "B", "C"),
  sand = c(95, 60, 20),
  silt = c(3, 30, 60),
  clay = c(2, 10, 20)
)

classify_texture(
  ssc,
  scheme = "gradistat",
  method = "rules",
  basis = "sand_silt_clay_no_gravel"
)
```

GRADISTAT rule classification returns `texture_class_id`, `texture_class`,
`classification_method`, `classification_status`, `ternary_basis`, `notes`,
and a ratio audit column. Input percentages must sum to approximately 100 for
the selected basis; grainsizeR does not silently normalize invalid sums.

`TEXTURAL GROUP` and `SEDIMENT NAME` are distinct GRADISTAT-style outputs.
Sediment-name composition can be requested during GRADISTAT classification with
`include_sediment_name = TRUE`, or added afterward with
`gs_gradistat_sediment_name()`. If dominant size-subclass columns are missing,
the textural group is returned as a partial sediment name.

The same supported bases can be drawn as texture ternary plots.
For the dry-sieve wide example, sample labels are suppressed to keep the
GRADISTAT gravel-sand-mud class labels readable. In the gravel-sand-mud basis,
the ternary plot places `Gravel` at the top, `Mud` at the lower-left apex, and
`Sand` at the lower-right apex. Gravel percentage guides and sand/mud ratio
guides are drawn as ternary annotations rather than Cartesian axes; unlike the
USDA display, these GRADISTAT apex labels remain at the vertices.

```{r}
plot_texture_ternary(
  head(gsm, 6),
  scheme = "gradistat",
  basis = "gravel_sand_mud",
  point_id = "sample_id",
  show_sample_labels = FALSE
)

plot_texture_ternary(
  ssc,
  scheme = "gradistat",
  basis = "sand_silt_clay_no_gravel",
  point_id = "sample_id"
)
```

Distribution, cumulative, fraction, USDA texture ternary, and GRADISTAT
texture ternary plots are R-native functional replacements, not Excel visual
clones.

`plot_texture_ternary()` is the preferred plotting name in new examples.
`plot_texture_triangle()` remains available for compatibility with existing
code.

## Output Columns

USDA rule classification returns the input rows with these classification
columns appended:

- `texture_class_id`
- `texture_class`
- `classification_method`
- `rule_status`
- `all_rule_matches`
- `rule_conflict`
- `rule_gap`

For valid, clean USDA inputs, `classification_method` is
`"usda_major_rules"` and `rule_status` is `"classified"`.

## User-Supplied Polygon Classification

Generic polygon classification remains available for user-provided texture
systems. The example below defines a synthetic full ternary region for
demonstration only; it is not an official texture system.

```{r}
polygons <- data.frame(
  scheme = "synthetic_ternary",
  class_id = "all",
  class_name = "Synthetic full ternary area",
  vertex_id = 1:3,
  left = c(100, 0, 0),
  right = c(0, 100, 0),
  top = c(0, 0, 100),
  left_component = "sand",
  right_component = "silt",
  top_component = "clay",
  reference_id = NA_character_,
  reference = NA_character_
)

polygons <- validate_texture_polygons(polygons)
```

```{r}
synthetic <- data.frame(
  sample_id = rep("A", 4),
  size_mm = c(2, 0.05, 0.002, 0.001),
  retained = c(10, 40, 30, 20)
)

synthetic_gs <- as_gsd_tbl(
  synthetic,
  sample_id,
  size_mm,
  retained,
  value_type = "percent"
)

classify_texture(
  synthetic_gs,
  texture_polygons = polygons,
  scheme = "synthetic_ternary",
  method = "polygon"
)
```

## Limitations

The USDA workflow covers only the 12 major USDA textural classes.
Sand-size modifier subclasses such as coarse sand, fine sand, very fine sand,
coarse sandy loam, fine sandy loam, and very fine sandy loam are not
implemented. They may be considered later as qualitative descriptor columns
for D50 or particle-size summaries.

No USDA polygon dataset or USDA rule-data object is bundled as runtime package
data.

GRADISTAT gravel-sand-mud and sand-silt-clay no-gravel ternary plotting is
implemented through `plot_texture_ternary()`. Full Excel visual parity is not
claimed.

CRAN readiness is not claimed in this vignette.

## Provenance and Validation

The public USDA classification path uses an internal rule helper validated
against USDA/NRCS validation examples and boundary cases. Public API tests
cover representative official, targeted, and boundary points.

Performance notes are informational. The USDA rule classifier is vectorized
and expected to be lighter than repeated polygon point-in-polygon checks for
this built-in USDA path, but formal benchmarking is not included.

The public GRADISTAT classification path uses independently written R decision
rules based on the documented GRADISTAT behavior. The workbook was used only as
a reference implementation, and VBA source code was not copied.
